Allocation Removal by Partial Evaluation in a Tracing JIT

### 2010-10-07 talk/pepm2011/paper.tex
rewrite to speak of "static objects" everywhere

The most important technique we use to achieve to optimize traces is a
straightforward application of partial evaluation. The partial evaluation
performs a form of escape analysis \cite{XXX} on the traces and make some
objects that are allocated in the trace \emph{static}\footnote{These objects are
called \emph{virtual} in Psyco \cite{rigo_representation-based_2004}.} which
means that they do not occur any more in the optimized trace. This technique is
informally described in Section~\ref{sec:virtuals}, a more formal description is
given in Section~\ref{sec:formal}.

@@

In the next section, we will see how this can be improved upon, using escape
analysis. XXX

@@

The creation of objects in category 1 is removed by the optimization described
in Section~\ref{sec:virtuals}. XXX

\section{Escape Analysis in a Tracing JIT} % XXX imprecise title
\label{sec:virtuals}

\subsection{Virtual Objects}

@@

XXX rewrite to acknowledge the existence of the formal stuff

This process is called \emph{escape analysis}. The escape analysis of
our tracing JIT works by using \emph{virtual objects}: The trace is walked from
beginning to end and whenever a \texttt{new} operation is seen, the operation is
removed and a shape description is constructed and associated with the variable
that would have stored the result of \texttt{new}. This result, a newly
allocated object, is called \emph{virtual}, because after the optimization is
finished, it does not need to be allocated at all. The shape description summarizes
the shape of this original object
and is used by the optimization to improve the trace. The shape describes
where the values that would be stored in the fields of the allocated objects
come from. Whenever the optimizer sees a \texttt{setfield} that writes into such
an object, that shape description is updated and the operation can be removed.
When the optimizer encounters a \texttt{getfield} from such an object, the result is read
from the shape description, and the operation is also removed.
Equivalently, a \texttt{guard\_class} on a variable that has a shape description can be removed as
well, because the shape description stores the class.

In the example from last section, the following operations would produce two
shape descriptions, and be completely removed from the optimized trace:

@@

The shape description associated with $p_{5}$ would know that it is an
\texttt{BoxedInteger}, and that the \texttt{intval} field contains $i_{4}$, the
associated with $p_{6}$ would know that its \texttt{intval} field contains the
constant -100.

@@

So far we have only described what happens when virtual objects are used in
operations that read and write their fields and in guards. When the virtual
object is used in
any other operation, it cannot stay virtual. For example, when a virtual object
is stored in a globally accessible place, the object needs to actually be
allocated, as it might live longer than one iteration of the loop. The actual
allocation is inserted just before the operation that requires the object to be
non-virtual.

This is what happens at the end of the trace in Figure~\ref{fig:unopt-trace}, when the \texttt{jump} operation
is hit. The arguments of the jump are at this point virtual objects. Before the
jump is emitted, they are \emph{forced}. This means that the optimizers produces code
that allocates a new object of the right type and sets its fields to the field
values that the virtual object has. This means that instead of the jump, the
following operations are emitted:

@@

Note how the operations for creating these two instances have been moved down the
trace. It looks like for these operations we actually didn't win much, because
the objects are still allocated at the end. However, the optimization was still
worthwhile even in this case, because some operations that have been performed
on the forced virtual objects have been removed (some \texttt{getfield} operations
and \texttt{guard\_class} operations).

@@

XXX Category 2 The optimization of
Section~\ref{sec:virtuals} deals with them too: the \texttt{new} that creates them and
the field accesses are deferred, until the point where the object escapes.

>>>>

The most important technique we use to achieve to optimize traces is a
straightforward application of partial evaluation. The partial evaluation
performs a form of escape analysis \cite{XXX} on the traces and make some
objects that are allocated in the trace \emph{static}\footnote{These objects are
called \emph{virtual} in Psyco \cite{rigo_representation-based_2004}.} which
means that they do not occur any more in the optimized trace. This technique is
informally described in Section~\ref{sec:statics}, a more formal description is
given in Section~\ref{sec:formal}.

@@

In the next section, we will see how this can be improved upon, using partial
evaluation. XXX

@@

The creation of objects in category 1 is removed by the optimization described
in Section~\ref{sec:statics}. XXX

\section{Allocation Removal in Traces}
\label{sec:statics}

\subsection{Static Objects}

@@

This is a process that is usually called \emph{escape analysis}. In this paper we will
perform escape analysis by using partial evaluation. The partial evalution is a
bit peculiar in that there are not actually any constant arguments to the trace,
but it is only used to optimized operations within a trace. XXX mention Prolog.

The partial evaluation works by walking the trace from beginning to end.
Whenever a \texttt{new} operation is seen, the operation is removed and a static
object is constructed and associated with the variable that would have stored
the result of \texttt{new}. The static object describes the shape of the
original object, \eg where the values that would be stored in the fields of the
allocated object come from, as well as the type of the object. Whenever the
optimizer sees a \texttt{setfield} that writes into such an object, that shape
description is updated and the operation can be removed, which means that the
operation was done at partial evaluation time. When the optimizer encounters a
\texttt{getfield} from such an object, the result is read from the shape
description, and the operation is also removed. Equivalently, a
\texttt{guard\_class} on a variable that has a shape description can be removed
as well, because the shape description stores the type.

In the example from last section, the following operations would produce two
static objects, and be completely removed from the optimized trace:

@@

The static object associated with $p_{5}$ would know that it is a
\texttt{BoxedInteger}, and that the \texttt{intval} field contains $i_{4}$, the
one associated with $p_{6}$ would know that its \texttt{intval} field contains
the constant -100.

@@

So far we have only described what happens when static objects are used in
operations that read and write their fields and in guards. When the static
object is used in any other operation, it cannot stay static. For example, when
a static object is stored in a globally accessible place, the object needs to
actually be allocated, as it might live longer than one iteration of the loop
and because the partial evaluator looses track of it. This means that the static
objects needs to be turned into a dynamic one, \ie lifted. This makes it
necessary to put operations into the residual code that actually allocate the
static object at runtime.

This is what happens at the end of the trace in Figure~\ref{fig:unopt-trace}, when the \texttt{jump} operation
is hit. The arguments of the jump are at this point static objects. Before the
jump is emitted, they are \emph{lifted}. This means that the optimizers produces code
that allocates a new object of the right type and sets its fields to the field
values that the static object has (if the static object points to other static
objects, those need to be lifted as well) This means that instead of the jump,
the following operations are emitted:

@@

Note how the operations for creating these two instances have been moved down the
trace. It looks like for these operations we actually didn't win much, because
the objects are still allocated at the end. However, the optimization was still
worthwhile even in this case, because some operations that have been performed
on the lifted static objects have been removed (some \texttt{getfield} operations
and \texttt{guard\_class} operations).

@@

XXX Category 2 The optimization of
Section~\ref{sec:statics} deals with them too: the \texttt{new} that creates them and
the field accesses are deferred, until the point where the object escapes.
